Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-25 omrp.tex
Streamlining Sec 3: OMRP
Mostly minor edits to make it read a bit better

Taking $n$ observations of the random variables $\xit$ and splitting them into $k$ batches of batch size $m$, define 
\begin{align} 
	\gb_j =\  & \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}),& \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         =\  & \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right],& \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress it from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that, in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but all the observations in that batch $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{m(j-1)+m}$ because $\xs_j$ is the optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ estimates are obtained, the overall mean of these estimates $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The sample variance is obtained as in (\ref{eq:var}) by $\varh{\gbb} = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in a one-sided CI $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varh{\gbb}} \right]$ that has a level of significance which is approximately $\alpha$.

@@

As before, let $m$ denote the batch size, $n$ the total sample size and $k = \lfloor \tfrac{n}{m} \rfloor$ be the number of nonoverlapping batches.  
In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the classical case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally overlapping case of Meketon and Schmeiser \citep{Meketon1984}.  
The sample mean of each batch estimator is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
where $\nb$ is the number of batches used given $n$, $m$ and $\gamma$.  
The sample variance estimator given in (\ref{eq:overlap_variance_def}) is changed to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
Note that with $\gamma=1$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:overlap_variance_def}) and with $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).

The amount of asymptotic variance reduction in this estimator depends on the asymptotic ratio of $\gamma/m$, which we denote by $\gammab$.  
For example, when $\gamma = 1$ (or $\gammab=0$; the maximally overlapping case), the variance is reduced to two-thirds ($66.67\%$) of the original nonoverlapping case, as given in (\ref{eq:asym_var}).  
When 75\% of observations overlap (i.e., when $\gamma = m/4$ or $\gammab = 1/4$), the variance is $33/48$th of the original ($68.75\%$), which is near-optimal.   
When only half of the observations overlap ($\gammab = 1/2$) the variance is 75\% of original.  
In general, from the spectral analysis given in Welch \citep{Welch1987}, the variance is reduced to
\begin{equation} \label{eq:var_reduct_formula}
	\gammab \left( 1 + 2 \sum_{j=1}^{\cogb-1} (1-j\gammab)^2 \right) = -\gammab\left( 1 - 2\gammab\cogb + \gammab^2\cogb^2 \right) + 2\gammab\cogb\left(1-\gammab\cogb\right) + \frac{2}{3}\gammab^3\cogb^3 + \frac{1}{3}\gammab^3\cogb
\end{equation}
of the nonoverlapping case.  
If $\ogb$ is an integer, the terms in parentheses vanish and (\ref{eq:var_reduct_formula}) simplifies to $\frac{2+\gammab^2}{3}$.

A $(1-\alpha)$-level approximate CI on the mean can be formed using variably overlapping batches by $\ybb \pm t_{d_{\gamma}(k-1),\alpha/2}$ $\sqrt{\vartg{\ybb}}$, where the degrees of freedom increase $d_{\gamma}$ is the multiplicative inverse of (\ref{eq:var_reduct_formula}) \citep{Welch1987}.  
If $\ogb$ is an integer, $d_{\gamma} = \frac{3}{2+\gammab^2}$ and for maximally overlapping batches, $d_{\gamma} = \frac{3}{2}$.

\subsection{Definition of Estimators for OMRP}
In order to apply overlapping batches to MRP, we need to keep track of solutions to sampling problems (SP$_m$) for each batch of size $m$.  
Toward this end, let $B(i)$ denote the set of batches $j \in \left\{1, 2, \dots, \nb \right\}$ observation $\xiti$ is used in, $i = 1, 2, \dots, n$.  
See Figure \ref{fig:overlap_nonint} for an example with $n=18$, $m = 6$ and $\gamma = 2$.  
Here, the first batch uses observations $\xit^1, \xit^2, \xit^3, \xit^4, \xit^5, \xit^6$, the second batch uses $\xit^3, \xit^4, \xit^5, \xit^6, \xit^7, \xit^8$, and so on.  
So, $B(1) = \{1\}$, $B(3) = \{1,2\}$, and $B(7)=\{2,3,4\}$.  
We use $\xs_j$ to denote an optimal solution to sampling problem (SP$_m$) formed using the $j$th batch; $\xs_j \in \argmin_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xit^{\gamma(j-1) + i})$, $j = 1, 2, \dots, \nb$.

>>>>

Taking $n$ observations of the random variables $\xit$ and splitting them into $k$ nonoverlapping batches of size $m$, define 
\begin{align} 
	\gb_j =\  & \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}),& \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         =\  & \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right],& \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress it from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that, in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but all the observations in that batch $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{m(j-1)+m}$ because $\xs_j$ is the optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ estimates are obtained, the overall mean of these estimates $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The sample variance is obtained as in (\ref{eq:var}) by $\varh{\gbb} = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in a one-sided CI $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varh{\gbb}} \right]$ that has a level of significance which is approximately $\alpha$.

@@

As before, let $m$ denote the batch size, $n$ the total sample size and $k = \lfloor \tfrac{n}{m} \rfloor$ be the number of nonoverlapping batches.  
In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the classical case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally overlapping case of \citet{Meketon1984}.  
The sample mean of each batch estimator is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
where $\nb$ is the number of batches used given $n$, $m$ and $\gamma$.  
The sample variance estimator given in (\ref{eq:overlap_variance_def}) is changed to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
Note that with $\gamma=1$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:overlap_variance_def}) and with $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).

The amount of asymptotic variance reduction in this estimator depends on the asymptotic ratio of $\gamma/m$, which we denote by $\gammab$.  
For example, when $\gamma = 1$ (or $\gammab=0$; the maximally overlapping case), the variance is reduced to two-thirds ($66.67\%$) of the original nonoverlapping case, as given in (\ref{eq:asym_var}).  
When 75\% of observations overlap (i.e., when $\gamma = m/4$ or $\gammab = 1/4$), the variance is $33/48$th of the original ($68.75\%$), which is near-optimal.   
When only half of the observations overlap ($\gammab = 1/2$) the variance is 75\% of original.  
In general, from the spectral analysis given in \citep{Welch1987}, the variance is reduced to
\begin{equation} \label{eq:var_reduct_formula}
	\gammab \left( 1 + 2 \sum_{j=1}^{\cogb-1} (1-j\gammab)^2 \right) %= -\gammab\left( 1 - 2\gammab\cogb + \gammab^2\cogb^2 \right) + 2\gammab\cogb\left(1-\gammab\cogb\right) + \frac{2}{3}\gammab^3\cogb^3 + \frac{1}{3}\gammab^3\cogb
\end{equation}
of the nonoverlapping case.  
If $\ogb$ is an integer, (\ref{eq:var_reduct_formula}) simplifies to $\frac{2+\gammab^2}{3}$.

A $(1-\alpha)$-level approximate CI on the mean can be formed using variably overlapping batches by $\ybb \pm t_{d_{\gamma}(k-1),\alpha/2}$ $\sqrt{\vartg{\ybb}}$, where the degrees of freedom increase $d_{\gamma}$ is the multiplicative inverse of (\ref{eq:var_reduct_formula}) \citep{Welch1987}.  
If $\ogb$ is an integer, $d_{\gamma} = \frac{3}{2+\gammab^2}$.  For maximally overlapping batches, the degrees of freedom is $d_{\gamma} = \frac{3}{2}$.

\subsection{Definition of Estimators for OMRP}
In order to apply overlapping batches to MRP, we need to keep track of solutions to sampling problems (SP$_m$) for each batch of size $m$.  
Toward this end, let $B(i)$ denote the set of batches $j \in \left\{1, 2, \dots, \nb \right\}$ observation $\xiti$ is used in, $i = 1, 2, \dots, n$.  
See Figure \ref{fig:overlap_nonint} for an example with $n=18$, $m = 6$ and $\gamma = 2$.  
Here, the first batch uses observations $\xit^1, \xit^2, \xit^3, \xit^4, \xit^5, \xit^6$, the second batch uses $\xit^3, \xit^4, \xit^5, \xit^6, \xit^7, \xit^8$, and so on.  
So, $B(1) = \{1\}$, $B(3) = \{1,2\}$, and $B(7)=\{2,3,4\}$.  
As before, we use $\xs_j$ to denote an optimal solution to sampling problem (SP$_m$) formed using the $j$th batch; $\xs_j \in \argmin_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xit^{\gamma(j-1) + i})$, $j = 1, 2, \dots, \nb$.
